Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-07-17 surfaces-write-up/surface-write-up.tex
some line breaks

\begin{example}
\label{eg:hyperplane}
Suppose $D = \frac{1}{2}H_0 - \frac{1}{3}H_1$ where $H_0 = V(x_0), H_1 = V(x_1)$ are two coordinate hyperplanes in $\bp^2$. Then, $R_D$ has generators in degree $2$ and $3$ which can be written as $u^2 \frac{x_1}{x_0}, u^3 \frac{x_1}{x_0}.$ In fact, for all degrees less than $5$ the elements of $R_D$ can all be expressed as rational functions in $x_0, x_1$. However, in degree $6$, there is the rational function $\frac{x_1^2 x_2}{x_0^3}.$ Since this involves $x_2$, it must be a generator.

This example generalizes slightly to any divisor of the form $D = \frac{1}{k}H_0 - \frac{1}{k+1}H_1 \in \di \bp^2 \otimes \bq,$ with $k \in \bn$, showing that there will always exist a generator in degree $a(a+1).$

This example further generalizes to the following situation: Suppose $D = \sum_{i=1}^{n} \frac{\num_i}{\den_i}D_i \in \di \bp^m \otimes \bq,$ where $\deg D_i = \bida_i$ and $\deg D = \frac{1}{\lcm_{1 \leq i \leq n}(\den_i \cdot \bida_i)}$. Then, if $D_i = V(f_i)$ where all $f_i$ can be written as a polynomial function in $x_0, \ldots, x_{m-1},$ we have $R_D$ always has a generator in degree $\lcm(\den_i \cdot \bida_i)$.
\end{example}

\begin{rem}
\label{rem:ghost-motivation}
Example ~\ref{eg:hyperplane} shows that the naive generalization of ~\cite[Theorem 8]{dorney:canonical} of generation in degree at most $\sum_{i=1}^{n}\ell_i$ cannot possibly hold. The reason for this is that the divisors may ``be contained in a hyperplane,'' i.e., be expressable as functions in $m$ of the $m+1$ variables on $\bp^m$. This problem can easily be circumvented by adding in ``ghost points.'' That is, we may add divisors of the form $0 \cdot H_i$ to $D$, and reorder so that if $D = \sum_{i=1}^{n}\alpha_i V(f_i)$, then $f_0, \ldots, f_n$.

@@

In this section, the main aim is to prove Theorem ~\ref{thm:}, bounding the degree of generators and relations of the canonical ring of a $\bq$-divisor on any hirzebruch surface.

>>>>

\begin{example}
\label{eg:hyperplane}
Suppose $D = \frac{1}{2}H_0 - \frac{1}{3}H_1$ where $H_0 = V(x_0),
H_1 = V(x_1)$ are two coordinate hyperplanes in $\bp^2$. Then, $R_D$
has generators in degree $2$ and $3$ which can be written as $u^2
\frac{x_1}{x_0}, u^3 \frac{x_1}{x_0}.$ In fact, for all degrees
less than $5$ the elements of $R_D$ can all be expressed as
rational functions in $x_0, x_1$. However, in degree $6$, there is
the rational function $\frac{x_1^2 x_2}{x_0^3}$. Since this
involves $x_2$, it must be a generator.

This example generalizes slightly to any divisor of the form $D =
\frac{1}{k}H_0 - \frac{1}{k+1}H_1 \in \di \bp^2 \otimes \bq,$ with
$k \in \bn$, showing that there will always exist a generator in
degree $a(a + 1)$.

This example further generalizes to the following situation:
Suppose $D = \sum_{i=1}^{n} \frac{\num_i}{\den_i}D_i \in \di \bp^m
\otimes \bq,$ where $\deg D_i = \bida_i$ and $\deg D = \frac{1}{\lcm
_{1 \leq i \leq n}(\den_i \cdot \bida_i)}$. Then, if $D_i = V(f_i)$
where all $f_i$ can be written as a polynomial function in $x_0,
\ldots, x_{m-1},$ we have $R_D$ always has a generator in degree
$\lcm(\den_i \cdot \bida_i)$.
\end{example}

\begin{rem}
\label{rem:ghost-motivation}
Example ~\ref{eg:hyperplane} shows that the naive generalization of
~\cite[Theorem 8]{dorney:canonical} of generation in degree at most
$\sum_{i=1}^{n}\ell_i$ cannot possibly hold. The reason for this is
that the divisors may ``be contained in a hyperplane,'' i.e.~ be
expressible as functions in $m$ of the $m+1$ variables on $\bp^m$.
This problem can easily be circumvented by adding in ``ghost
points.'' That is, we may add divisors of the form $0 \cdot H_i$ to
$D$, and reorder so that if $D = \sum_{i=1}^{n}\alpha_i V(f_i)$,
then $f_0, \ldots, f_n$.

@@

In this section, the main aim is to prove Theorem ~\ref{thm:}, bounding the degree of generators and relations of the canonical ring of a $\bq$-divisor on any Hirzebruch surface.
